Explore tumor cell classification with CellAssign

### 2024-05-22 analyses/cell-type-ewings/template_notebooks/02-cellassign.Rmd
account for missing tumor markers and no cell assign results

```{r}
# function to get assigned cell type based on max prediction returned by CellAssign 
get_celltype_assignment <- function(predictions){
  # get individual cell type assignments
  # those with the max prediction 
  celltype_assignments <- predictions |>
    tidyr::pivot_longer(
      !barcode,
      names_to = "celltype",
      values_to = "score"
    ) |>
    dplyr::group_by(barcode) |>
    dplyr::slice_max(score, n = 1) |>
    dplyr::ungroup()

@@

@@

# create a table of assignments 
table(tumor_celltype_assignments$celltype)
```

```{r}
# get annotations for heatmaps
annotation_df <- data.frame(
  marker_gene_classification = classifications_df$marker_gene_classification,
  row.names = classifications_df$barcodes
)

# heatmap of prediction scores 
tumor_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

```{r}
ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = cellassign_annotations)) +
  geom_point(size = 0.5, alpha = 0.5) + 
  theme_bw()
```

```{r}
# prep annotations for confusion matrix
celltype_df <- celltype_df |> 
  dplyr::mutate(cellassign_annotations = ifelse(
    # make sure annotations match marker gene annotations
    cellassign_annotations == "tumor", "Tumor", "Normal"
  )) |> 
  # make tumor the positive class
  dplyr::mutate(
    cellassign_annotations = forcats::fct_relevel(cellassign_annotations, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  )

@@

```{r}
# get marker gene expression 
marker_gene_exp <- logcounts(sce[tumor_markers_df$ensembl_gene_id, ]) |>
  as.matrix() |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes")

@@

table(visser_celltype_assignments$celltype)
```

```{r}
# heatmap of prediction scores 
visser_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

# create a table of assignments 
table(panglao_celltype_assignments$celltype)
```

```{r}
# heatmap of prediction scores 
panglao_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

>>>>

```{r}
# function to get assigned cell type based on max prediction returned by CellAssign 
get_celltype_assignment <- function(predictions){

  # if no predictions, assign every cell to NA
  if(length(colnames(predictions)) == 1){
    celltype_assignments <- data.frame(
      barcode = predictions$barcode,
      celltype = NA
    )
  } else {
    # get individual cell type assignments
    # those with the max prediction 
    celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "score"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(score, n = 1) |>
      dplyr::ungroup() 
  }

@@

```{r}
# check if marker gene annotations are present 
if(all(is.na(classifications_df$marker_gene_classification))){
  has_marker_gene <- FALSE
} else {
  has_marker_gene <- TRUE 
}
```

@@

if(all(is.na(tumor_celltype_assignments$celltype))){
  has_tumor_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(tumor_celltype_assignments$celltype) 
  has_tumor_assignments <- TRUE
}

```{r}
# define annotations to use in all heatmaps
if(has_marker_gene){
  # get annotations for heatmaps
  annotation_df <- data.frame(
    marker_gene_classification = classifications_df$marker_gene_classification,
    row.names = classifications_df$barcodes
  ) 
} else {
  annotation_df <- NULL
}
```

```{r, eval=has_tumor_assignments}
# heatmap of prediction scores 
tumor_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

```{r, eval=has_tumor_assignments}
ggplot(celltype_df, aes(x = UMAP1, y = UMAP2, color = cellassign_annotations)) +
  geom_point(size = 0.5, alpha = 0.5) + 
  theme_bw()
```

```{r, eval=has_tumor_assignments}
# prep annotations for confusion matrix
celltype_df <- celltype_df |> 
  dplyr::mutate(cellassign_annotations = ifelse(
    # make sure annotations match marker gene annotations
    cellassign_annotations == "tumor", "Tumor", "Normal"
  )) |> 
  # make tumor the positive class
  dplyr::mutate(
    cellassign_annotations = forcats::fct_relevel(cellassign_annotations, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  )

@@

```{r, eval=has_marker_gene}
# get marker gene expression 
marker_gene_exp <- logcounts(sce[tumor_markers_df$ensembl_gene_id, ]) |>
  as.matrix() |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes")

@@

if(all(is.na(visser_celltype_assignments$celltype))){
  has_visser_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(visser_celltype_assignments$celltype) 
  has_visser_assignments <- TRUE
}
```

```{r, eval=has_visser_assignments}
# heatmap of prediction scores 
visser_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

if(all(is.na(panglao_celltype_assignments$celltype))){
  has_panglao_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(panglao_celltype_assignments$celltype) 
  has_panglao_assignments <- TRUE
}
```

```{r, eval=has_panglao_assignments}
# heatmap of prediction scores 
panglao_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```
